Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-09-02 paper/submissions/science/main.tex

% Figures
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\def\figdir{../Figs}

@@

% Refs
%\usepackage{scicite}
\usepackage[style=science,
					backend=biber,
					sortcites=true,
					autocite=superscript
]{biblatex}

@@

%
% ABSTRACT
%
\newpage
\begin{sciabstract}
	Human mobility drives major societal phenomena including epidemics, economies, and innovation.
	Historically, mobility was constrained by geographic distance, however, in the globalizing world, language, culture, and history are increasingly important.
	Here, we show a mathematical equivalence between \textit{word2vec} model and the gravity model of mobility and demonstrate that, by using three human trajectory datasets, \textit{word2vec} encodes nuanced relationships between locations into a vector-space, providing a measure of effective distance.
	Focusing on the case of scientific mobility, we show that embeddings uncover cultural, linguistic, and hierarchical relationships at multiple levels of granularity.
	Connecting neural embeddings to the gravity model opens up new avenues for the study of mobility.
\end{sciabstract}

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Although geographic distance has historically constrained human movements, it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, the decreasing relevance of geography calls for alternative ways of conceptualizing ``distance''~\autocite{boschma2005proximity, brown1970migration, kim2018functional}.

Yet, functional distances are often low-resolution, computed at the level of countries rather than regions, cities, or organizations, and have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.
Low dimensional distance alone cannot represent the multitude of inter-related factors that drive mobility.
Networks offer a solution to representing many dimensions of mobility, yet edges only encode simple relationships between connected entities.
Capturing the complexity of mobility requires moving beyond simple functional distances and networks, to learning high-dimensional landscapes of mobility that incorporate many facets of mobility into a single fine-grained and continuous representation.

Here, we demonstrate that the \textit{word2vec} model (Skip-Gram Negative Sampling) is equivalent to the gravity law of mobility, a fundamental mobility framework. 
We then empirically test the resulting representation by its ability to derive functional effective distances between from trajectories.
After validating that it accurately represents real-world data, we demonstrate \textit{word2vec}'s capacity for encoding rich information relating to geography, culture, language, and even prestige, at multiple scales of analysis. 
We embed trajectories from three datasets: U.S. passenger flight itinerary records, Korean accommodation reservations, and a dataset of scientists' career mobility between organizations captured in bibliometric records (Detailed descriptions are available in the Methods).

We focus in more detail on scientific mobility due to its richness and importance.
Scientific mobility---which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, morgan2018prestige}---is not only an important topic in the Science of Science but also ideal for our study thanks to its well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{clauset2015hierarchy, deville2014career}.
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, deville2014career}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility. Trajectories of scientific mobility are constructed using more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:gravity_performance}a).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, sugimoto2017mostimpact} (see \SI).

Here, we study the skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}, networks~\autocite{grover2016node2vec, linzhuo2020hyperbolic} and trajectories\autocite{liu2016exploring, feng2017poi2vec, yao2018representing, cao2019habit2vec, crivellari2019motion, solomon2018predict}.
It works under the notion that a good representation should facilitate prediction, learning a mapping between words that can predict a target word based on its context (surrounding words).
The model is also computationally efficient, easy to use, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis, hashimoto2016word}.
As a result, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory.

Our study builds on the gravity model framework~\autocite{zipf1946gravity}.
This is a widely-used, fundamental mobility model~\autocite{curiel2018citygravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity} that connects the \textit{expected flux}, $\hat{T}_{ij}$, between locations based on their populations and distance:

@@

%
% Wordvec and the gravity model
\subsection*{word2vec and the gravity model}

We first demonstrate the mathematical equivalence between the SGNS model and the gravity model.
The {\it word2vec} model takes a location trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), as input.
A target location $a_t=i$ is considered to have a context location $a_{t'}=j$ that appears in the previous or subsequent $w$ locations in the trajectory, \ie  $j \in [a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}]$.
	{\it word2vec} learns an embedding by estimating the probability that location $i$ has context $j$:
\begin{align}
	P\left(j \given i \right):= \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng}
\end{align}
where the denominator $Z_i=\sum_{j' \in \mathcal{A}} \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
Although {\it word2vec} generates two embedding vectors $\bm{v_i}$ and $\bm{u}_i$---referred to as the in-vector and out-vector, respectively---we follow convention to use the in-vector $\bm{v}_i$ as an embedding of location $i$.

When $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), the flow can be written as
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} \propto P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(j)$ is the fraction of $j$ in the data.
In general, calculating $Z_i$ is computationally expensive and there are two common approximations: hierarchical softmax \autocite{morin2005hierarchical} and negative sampling \autocite{mikolov2013word2vec}.
Due to its simplicity and performance, negative sampling is the most widely used strategy, which we also adopt in our study.

@@

In other words, \emph{word2vec with the skip-gram negative sampling is equivalent to the gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model describes mobility flows from the given mass and locations, {\it word2vec} finds locations that best explain the given mobility flow. 
\js{Also, our finding is aligned with a perspective on word embedding methods as a metric recovery~\autocite{hashimoto2016word}, which recover the underlying metric of the semantic manifold not for a particular task.}

Given their equivalence, we expect that the \textit{word2vec} embedding can encode effective functional distances between locations according to the gravity law. 
Here, we test this notion using three human mobility datasets, showing that \textit{word2vec}  consistently offers a better representation of actual mobility flows than baselines methods.
In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$), which has traditionally been used to quantify distance for the gravity model.
Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d).
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).

%
% Figure - Example of mobility sentences
%
%
% Figure - Gravity law performance
%
\begin{figure}[!htp]
	\centering
	\includegraphics[width=0.9\textwidth]{\figdir/GravityLawFig/gravity_performance_new.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts mobility across three distinct human trajectory datasets. }
		\textbf{a.} A unique identifier is assigned to each organization and they are assembled into an affiliation trajectory ordered by year of publication (top). If an author lists multiple organization affiliations within the same year,  we shuffle the order within that year in each training iteration (bottom, see \SI).		
		\textbf{b.} Neural embeddings outperform baselines for scientific mobility.
The correlation between the flux and baseline distances $R^{2}$ (left), and the  root-mean-squared error between the actual and predicted values (right). The asterisk indicates the top-performing method.
		\textbf{c.} Embedding distance (top) better explains the expected flux of passengers between U.S. airports than does geographic distance (bottom).
		The red line is the line of the best fit.
		%Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted for the mean flux in each bin.
		Correlation is calculated on the data in the log-log scale ($p < 0.0001$ across all fits).
		%The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{d.} Predictions of flux between airport pairs made using embedding distance (top) outperform those made using geographic distance (bottom).
		Box-plots show the distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$.
		%a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
	Embedding distance consistently produce powerful functional distance compare to geographic distance for Korean accommodation reservations (\textbf{e,f}), and global scientific mobility (\textbf{g,h})}
	\label{fig:gravity_performance}
\end{figure}

The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores calculated from the underlying mobility network (Fig.~\ref{fig:gravity_performance}b, Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
The embedding distance derived from neural embedding also explains more of the flux and better predicts mobility flows than simpler embedding baselines, such as distances derived from a singular-value decomposition and a Laplacian Eigenmap embedding~\autocite{belkin2003laplacian} of the underlying location co-occurrence matrix, Levy's symmetric \textit{word2vec}\autocite{levy2014neural}, and even direct optimization of the gravity model (Fig.\ref{fig:gravity_performance}b and Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
In sum, our results demonstrate that, consistently and efficiently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

@@

\subsection*{U.S. flight itinerary data}
We source U.S. airport itinerary data from the Origin and Destination Survey (DB1B), provided by the Bureau of Transportation Statistics at the United States Department of Transportation.
DB1B is a sample of 10 percent of domestic airline tickets between 1993 and 2020, comprising 307,760,841 passenger itineraries between 828 U.S. airports.
A trajectory is constructed for each passenger flight itinerary, forming an ordered sequence of unique identifiers of the origin and destination airports.
Each itinerary is associated with a trajectory of airports including the origin, destination, and intermediary stops.

\subsection*{Korean accommodation reservation data}
We source Korean accommodation reservation data from collaboration with Goodchoice Company LTD.
The data contains customer-level reservation trajectories spanning the period of August 2018 through July 2020 and comprising 1,038 unique accommodation locations in Seoul, South Korea.
A trajectory is constructed for each customer, containing the ordered sequences of accommodations they reserved over time.

@@

%% Mobility sentences
We embed trajectories by treating them analogously to sentences and locations analogously to words.
For U.S. airport itinerary, trajectories are formed from the flight itineraries of individual passenger, in which airports correspond to unique identifiers.
In the case of Korean accommodation reservations, trajectories comprise a sequence of accommodations reserved over a customer's history.
For scientific mobility, an``affiliation trajectories'' is constructed for each mobile author, which is built by concatenating together their ordered list of unique organization identifiers, as demonstrated in Fig.~\ref{fig:gravity_performance}a (top).
In more complex cases, such as listing multiple affiliations on the same paper or publishing with different affiliations on multiple publications in the same year, the order is randomized within that year, as shown in Fig.~\ref{fig:gravity_performance}a (bottom).

@@

\subsection*{Element-centric clustering similarity}

Element-centric clustering similarity \autocite{gates2019element} is a similarity measure that can produce disjoint, overlapping, and hierarchically-structured clusterings. 
Element-centric clustering similarity captures cluster-induced relationships between elements through a cluster affiliation graph where one vertex set is the original element $V=\{v_1, ... v_N\}$ and the other corresponds to the cluster $C=\{c_1, ... c_M\}$ as a bipartite graph $\mathcal{B} (V \cup C, \mathcal{R})$. 
An undirected edge $a_{i\beta} \in \mathcal{R}$ denotes element $v_i$ as a member of cluster $c_\beta$. 
For hierarchically structured clustering, each cluster $c_\beta$ is assigned a hierarchical level $l_{\beta} \in [0,1]$ by re-scaling the dendrogram according to the maximum path length from its roots. 
The weight of the cluster affiliation edge is given by the hierarchy weight function $a_{i\beta}  = e^{rl_{\beta}}$ with scaling parameter $r$ which determines the relative importance of membership at different levels of hierarchy. 
Here, lower $r$ focuses on the higher levels of the hierarchy, and lower levels are treated as a refinement of the higher level. 
Larger $r$ emphasizes lower-level cluster structure, and higher levels of the hierarchy are seen as an aggregation of the lower-level structure. 
If $r=0$, every cluster is treated equally.

The cluster affiliation graph is projected onto a cluster-induced element graph which is a weighted, directed graph summarizing the relationship induced by common cluster memberships. 
In the cluster-induced element graph, each edge  between element $v_i$ and $v_j$ has weight $w_{ij} = \sum_{\gamma} \frac{a_{i\gamma}a_{j\gamma}}{\sum_{\kappa}a_{i\kappa} {\sum_{m}a_{m\gamma}}}$. 
Given a cluster-induces element graph with weighted matrix $W$, the personalized PageRank vector $p_i$  is used as membership-aware similarity between element $i$ and other elements in the graph. 
Then, the element-wise similarity of an element $v_i$ in two clusters  $\mathcal{A}$ and $\mathcal{B}$ is calculated with $S_i(\mathcal{A}, \mathcal{B}) = 1 - L_1(p_i^{\mathcal{A}},p_j^{\mathcal{B}})$, and the final element-centric similarity of two clustering $\mathcal{A}$ and $\mathcal{B}$ is found as the average of the element-wise similarities, $S(\mathcal{A}, \mathcal{B}) =\frac{1}{N}\sum_{i=1}^{N} S_i(\mathcal{A}, \mathcal{B})$.

>>>>

% Figures
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\def\figdir{../../Figs}

@@

% Refs
%\usepackage{scicite}
\usepackage[style=nature,
					backend=biber,
					sortcites=true,
					autocite=superscript
]{biblatex}

@@

%
% ABSTRACT
%
\newpage
\begin{sciabstract}
	Human mobility drives major societal phenomena including epidemics, economies, and innovation.
	Historically, mobility was constrained by geographic distance, however, in the globalizing world, language, culture, and history are increasingly important.
	We propose using the neural embedding model \textit{word2vec} for studying mobility and capturing its complexity.
    \textit{Word2ec} is shown to be mathematically equivalent to the gravity model of mobility, and using using three human trajectory datasets, we demonstrate that it encodes nuanced relationships between locations into a vector-space, providing a measure of effective distance that outperforms baselines. 
	Focusing on the case of scientific mobility, we show that embeddings uncover cultural, linguistic, and hierarchical relationships at multiple levels of granularity.
	Connecting neural embeddings to the gravity model opens up new avenues for the study of mobility.
\end{sciabstract}

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Although geographic distance has historically constrained human movements, it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, there have been attempts to define alternative notions of distance, or functional distances~\autocite{boschma2005proximity, brown1970migration, kim2018functional}, from real-world data or \textit{a priori} relationships between geographic entities.

Yet, functional distances are often low-resolution, computed at the level of countries rather than regions, cities, or organizations, and have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.
Low dimensional distance alone cannot represent the multitude of inter-related factors that drive mobility.
Networks offer a solution to representing many dimensions of mobility, yet edges only encode simple relationships between connected entities.
Capturing the complexity of mobility requires moving beyond simple functional distances and networks, to learning high-dimensional landscapes of mobility that incorporate the many facets of mobility into a single fine-grained and continuous representation.

Here, we apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense and continuous vector-space representation, from which we can not only derive a meaningful functional distance between locations but also probe relationships based on culture, language, and even prestige along with the geographic relationship.
We embed trajectories from three massive datasets: U.S. passenger flight itinerary records, Korean accommodation reservations, and a dataset of scientists' career mobility between organizations captured in bibliometric records (Detailed descriptions are available in the Methods).

The flight itinerary data, from the Airline Origin and Destination Survey, consists of records of more than 300 million itineraries between 1993 and 2020 documenting domestic flights between 828 airports in the United States.
A trajectory is constructed for each passenger flight itinerary, forming an ordered sequence of unique identifiers of the origin and destination airports.
The Korean accommodation reservations consist of customer reservation histories across 2018 and 2020 for 1,038 unique accommodation locations in Seoul, South Korea.
A trajectory is constructed for each customer, containing the ordered sequences of accommodations they reserved over time.
Finally, we use scientific mobility data that captures the affiliation trajectories of nearly 3 million scientists across ten years.
We focus in more detail on scientific mobility due to its richness and importance.
Scientific mobility---which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, morgan2018prestige}---is not only an important topic in the Science of Science but also ideal for our study thanks to its well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{clauset2015hierarchy, deville2014career}.
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, deville2014career}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

Trajectories of scientific mobility are constructed using more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, sugimoto2017mostimpact} (see \SI).

A vector-space embedding of locations (airports, accommodations, and organizations) is learned by using trajectories as input to the standard with skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}, networks~\autocite{grover2016node2vec, linzhuo2020hyperbolic} and trajectories\autocite{liu2016exploring, feng2017poi2vec, yao2018representing, cao2019habit2vec, crivellari2019motion, solomon2018predict}.
It works under the notion that a good representation should facilitate prediction, learning a mapping between words that can predict a target word based on its context (surrounding words).
The model is also computationally efficient, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis}.
As a result, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory.

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
	\centering
	\includegraphics[width=0.5\textwidth]{\figdir/mobility_example_combined.pdf}
	\caption{
		\textbf{Construction of affiliation trajectories from publication records.}
		\textbf{a.}
		An author published five papers across five time periods, with only one affiliation listed in the byline of each paper.
		A unique identifier is assigned to each organization and they are assembled into an affiliation trajectory ordered by year of publication.
		\textbf{b.}
		If an author lists multiple organization affiliations within the same year, then organization IDs within that year are placed in random order in each training iteration of the \textit{word2vec} model (for more detail, see \SI).
	}
	\label{fig:methods:mobility_sentence}
\end{figure}

To validate our approach, we evaluate the quality of vector representations based on their performance in predicting real-world mobility flows using the gravity model framework~\autocite{zipf1946gravity}.
The gravity model is a widely used mobility model~\autocite{curiel2018citygravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity} that connects the \textit{expected flux}, $\hat{T}_{ij}$, between locations based on their populations and distance:

@@

We show that the embedding distance better predicts actual mobility flows than the geographic distance across three disparate datasets.
In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$).
Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d).
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores calculated from the underlying mobility network (Fig.~\ref{fig:supp:distancemetrics}, Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
The embedding distance derived from neural embedding also explains more of the flux and better predicts mobility flows than simpler embedding baselines, such as distances derived from a singular-value decomposition and a Laplacian Eigenmap embedding~\autocite{belkin2003laplacian} of the underlying location co-occurrence matrix, Levy's symmetric \textit{word2vec}\autocite{levy2014neural}, and even direct optimization of the gravity model (Fig.\ref{fig:supp:distancemetrics} and Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
In sum, our results demonstrate that, consistently and efficiently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts mobility in flights, accommodation reservations, and global scientific mobility.}
		\textbf{a.} Embedding distance (cosine distance between organization vectors, top) better explains the expected flux of passengers between U.S. airports than does geographic distance (bottom).
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted for the mean flux in each bin based on a normal distribution.
		Correlation is calculated on the data in the log-log scale ($p < 0.0001$ across all fits).
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{b.} Predictions of flux between airport pairs made using embedding distance (top) outperform those made using geographic distance (bottom).
		Box-plots show the distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		Results are consistent in the case of scientific mobility.
		For Korean accommodation reservations, embedding distance better explains the expected flux than does geographic distance (\textbf{c}), and produces better predictions (\textbf{d}).
		Similarly, in the case of global scientific mobility, embedding distance explains the expected flux between organizations (\textbf{e)} and allows for better predictions (\textbf{f}) than geographic distance.
	}
	\label{fig:gravity_performance}
\end{figure}

\subsection*{word2vec and the gravity model}

Why does \textit{word2vec} with the skip-gram with negative sampling model (SGNS) work so well to model mobility?
The reason for that is the mathematical equivalence between the SGNS model and the gravity model which we demonstrate below.

The {\it word2vec} model takes a location trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), as input.
A target location $a_t=i$ is considered to have a context location $a_{t'}=j$ that appears in the previous or subsequent $w$ locations in the trajectory, \ie  $j \in [a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}]$.
	{\it word2vec} learns an embedding by estimating the probability that location $i$ has context $j$:
\begin{align}
	P\left(j \given i \right):= \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i}, \label{eq:prob_w2v_ng}
\end{align}
where the denominator $Z_i=\sum_{j' \in \mathcal{A}} \exp(\bm{u}_{j'} \cdot \bm{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations.
Although {\it word2vec} generates two embedding vectors $\bm{v_i}$ and $\bm{u}_i$---referred to as the in-vector and out-vector, respectively---we follow convention to use the in-vector $\bm{v}_i$ as an embedding of location $i$.

Our experiments show that {\it word2vec} best explains real-world mobility flow when $w=1$ (Fig.~\ref{fig:supp:hyperparameter}), with the flow predicted by
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} \propto P(i) P(j\given i) = \frac{P(i)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $P(j)$ is the fraction of $j$ in the data.
In general, calculating $Z_i$ is computationally expensive and there are two common approximations: hierarchical softmax \autocite{morin2005hierarchical} and negative sampling \autocite{mikolov2013word2vec}.
Due to its simplicity and performance, negative sampling is the most widely used strategy, which we also adopt in our study.

@@

In other words, \emph{word2vec with the skip-gram negative sampling is equivalent to the gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model predicts mobility flows from the given mass and locations, {\it word2vec} finds locations that best explain the given mobility flow.
Therefore, the embedding distance is inherently tied to the mobility flow, and hence, has greater predictive power than the geographic distance and other baselines.

@@

\subsection*{U.S. flight itinerary data}
We source U.S. airport itinerary data from the Origin and Destination Survey (DB1B), provided by the Bureau of Transportation Statistics at the United States Department of Transportation.
DB1B is a sample of 10 percent of domestic airline tickets between 1993 and 2020, comprising 307,760,841 passenger itineraries between 828 U.S. airports.
Each itinerary is associated with a trajectory of airports including the origin, destination, and intermediary stops.

\subsection*{Korean accommodation reservation data}
We source Korean accommodation reservation data from collaboration with Goodchoice Company LTD.
The data contains customer-level reservation trajectories spanning the period of August 2018 through July 2020 and comprising 1,038 unique accommodation locations in Seoul, South Korea.

@@

%% Mobility sentences
We embed trajectories by treating them analogously to sentences and locations analogously to words.
For U.S. airport itinerary, trajectories are formed from the flight itineraries of individual passenger, in which airports correspond to unique identifiers.
In the case of Korean accommodation reservations, trajectories comprise a sequence of accommodations reserved over a customer's history.
For scientific mobility, an``affiliation trajectories'' is constructed for each mobile author, which is built by concatenating together their ordered list of unique organization identifiers, as demonstrated in Fig.~\ref{fig:methods:mobility_sentence}a.
In more complex cases, such as listing multiple affiliations on the same paper or publishing with different affiliations on multiple publications in the same year, the order is randomized within that year, as shown in Fig.~\ref{fig:methods:mobility_sentence}b.

@@
